---
title: "Phase2_Kenny_B"
author: "Kenneth Burlingham"
date: "2025-11-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load libraries}
library(tidyverse)
```

Just want a way to look at the list of columns

```{r load data}
merged_data<-read.csv("merged_data.csv")
merged_data$SOVI_RATNG<-as.factor(merged_data$SOVI_RATNG) #cleaning
nri_cols<-names(merged_data)

length(unique(merged_data$STCOFIPS))==nrow(merged_data)

```

All counties with avalanche 

```{r check out useless columns}

col_summary <- merged_data |> 
  summarise(across(everything(), ~ case_when(
    all(is.na(.x)) ~ "all_na",
    n_distinct(.x, na.rm = TRUE) == 1 ~ "one_value",
    TRUE ~ "ok"
  ))) |> 
  pivot_longer(everything(), names_to = "column", values_to = "status")

# Separate results
all_na_cols <- col_summary |> filter(status == "all_na") |> pull(column)
one_value_cols <- col_summary |> filter(status == "one_value") |> pull(column)

list(
  entirely_na = all_na_cols,
  one_unique_value = one_value_cols
)

mean(is.na(merged_data$CFLD_EVNTS))
mean(is.na(merged_data$ERQK_EVNTS))
mean(is.na(merged_data$WFIR_EVNTS))

table(merged_data$AVLN_EXPB) #208 counties with building avalanche dmg
table(merged_data$AVLN_EXPP) 
table(merged_data$AVLN_EXPPE) #population equivalence
table(merged_data$AVLN_EXPT)

```

Columns counting events for coastal flooding, earthquakes and wildfire - completely empty
Columns counting avalanche exposure is only one value for 208 counties

```{r counties with avalanche exposure}
merged_data[which(!is.na(merged_data$AVLN_EXPB)),]%>%select(COUNTY,STATE)
```

Loss types are all skewed

```{r check skewness}
hist(merged_data$EAL_VALT)
hist(log(merged_data$EAL_VALT))

merged_data$log_EAL_VALT<-log(merged_data$EAL_VALT) #log since skewed
summary(merged_data$EAL_VALT)

#all skewed
hist(merged_data$EAL_VALA)
hist(merged_data$EAL_VALB)
hist(merged_data$EAL_VALPE)

summary(merged_data$EAL_VALA)
summary(merged_data$EAL_VALB)
summary(merged_data$EAL_VALPE)

merged_data$log_EAL_VALA<-log(merged_data$EAL_VALA+1)
merged_data$log_EAL_VALB<-log(merged_data$EAL_VALB)
merged_data$log_EAL_VALPE<-log(merged_data$EAL_VALPE)


```

```{r Exploring total expect loss}
merged_data%>%select(EAL_VALT,STATE,COUNTY)%>%arrange(desc(EAL_VALT))%>%head(10)

merged_data[which.max(merged_data$EAL_VALT),c("STATE","COUNTY")] #LA is the highest EAL_VALT

```

Wanted to check where there isn't any SOVI data, basically just territories of the US.

```{r SOVI exploration}
merged_data[which(merged_data$SOVI_RATNG=="Data Unavailable"),]%>%select(STATE,COUNTY,SOVI_SCORE)
#missing SOVI rating, American Samoa, Guam, Northern Mariana Islands,Virgin Islands
#Also missing SOVI Score, obviously

summary(merged_data$SOVI_SCORE)
hist(merged_data$SOVI_SCORE)

merged_data_s<-filter(merged_data,SOVI_RATNG!="Data Unavailable")

```

Created "merged_data_s", removing the ten counties from territories that are missing SOVI data.

```{r categorical sovi ratings explaining total loss}
table(merged_data$SOVI_RATNG)

#categorical sovi - loses too much information
sovi_cat_model<-lm(log_EAL_VALT~SOVI_RATNG,
                    data=merged_data)

summary(sovi_cat_model)

#trying categorical one more time with specific loss type
sovi_cat_ag<-lm(log_EAL_VALA~SOVI_RATNG,
            data=merged_data)
summary(sovi_cat_ag)
#holy shit?

sovi_cat_pe<-lm(log_EAL_VALPE~SOVI_RATNG,
                data=merged_data)
summary(sovi_cat_pe)

```

```{r numeric sovi explaining loss types}
#numeric sovi with total
sovi_num_model<-lm(log_EAL_VALT~SOVI_SCORE,
                   data=merged_data)
summary(sovi_num_model)

#numeric sovi with ag
sovi_ag<-lm(log_EAL_VALA~SOVI_SCORE,
            data=merged_data)
summary(sovi_ag)

#num sovi with human loss
sovi_pe<-lm(log_EAL_VALPE~SOVI_SCORE,
            data=merged_data)
summary(sovi_pe)

#num sovi with bldg loss
sovi_bld<-lm(log_EAL_VALB~SOVI_SCORE,
             data=merged_data)
summary(sovi_bld)
```

SOVI is not significant when converted to categorical, too much information is lost

Ag~SOVI coeff: A one point increase in SOVI is associated with a 1% decrease in agricultural damage
Pe~SOVI: 0.9% increase in person damage
Bld~SOVI: 1.65% increase in building damage (largest coefficient and R^2)


```{r plots}
ggplot(merged_data, aes(x = SOVI_SCORE, y = log_EAL_VALT)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "log(EAL_VALT) vs SOVI_SCORE",
    x = "SOVI Score",
    y = "Log of EAL_VALT"
  ) +
  theme_minimal()
```

The relationship between log losses + SOVI looks linear

```{r loss explaining sovi scores}
ag_loss<-lm(SOVI_SCORE~log_EAL_VALA,
            data=merged_data)
summary(ag_loss)

pe_loss<-lm(SOVI_SCORE~log_EAL_VALPE,
            data=merged_data)
summary(pe_loss)

bld_loss<-lm(SOVI_SCORE~log_EAL_VALB,
            data=merged_data)
summary(bld_loss)
```

Ag Loss: 1% increase in log ag loss is associated with an decrease of 0.009 sovi points
Pe loss: 1% increase in log pe loss > increase of 0.04 sovi points
Bld loss: 1% increase in log bld loss > increase of 0.042 sovi points
